---
title: "WP2 DeclareDesign Power Example"
format:
  typst:
    papersize: a4
    margins: 1in
execute:
  echo: true
  freeze: auto
  warning: false
  message: false
---

This document is the standalone version of the WP2 power example. It replaces
the separate R script and keeps the analysis, explanation, and executable code
in one place.

## Purpose

WP2 is a survey-embedded communication experiment among Danish citizens living
in coastal or high flood-risk areas. The planned panel has 4,000 respondents
interviewed twice during the year. Half are assigned to communication
treatments featuring facts and policy information framed around themes such as
community, solidarity, identities, flooding, climate adaptation, and shared
responsibility. The other half are assigned to control.

The relevant outcomes are support for climate policy and democratic attitudes:
trust in democracy, civil society, and local communities. The main goal of this
example is to show how a DeclareDesign power analysis can be built around those
assumptions.

## Effect-Size Prior

The example uses standardized treatment effects because the final WP2 outcome
scale is not fixed. It also creates a derived 1-5 observed index so the same
design can be inspected on a Likert-style scale.

The main simulation grid is:

```{r}
#| label: tau-preview
tau_grid <- c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30)
```

The reading of the three background papers was:

- Ballew et al. supports a conservative range around `0.10-0.25 SD` for
  communication-relevant predictors, but those are observational associations,
  not randomized treatment effects.
- Sinclair et al. supports a planning range around `0.10-0.30 SD`, with
  `0.20 SD` as a reasonable intervention anchor for intention-type outcomes.
- Goldwert et al. also supports `0.10`, `0.20`, and `0.30 SD` sensitivity
  cases, while warning that effects vary substantially by outcome, ideology,
  attrition, and multiple-testing choices.

The interpretation should therefore be cautious: `0.20 SD` is a defensible
main example, `0.10 SD` is a modest-effect case, and `0.30 SD` is optimistic.
The `0.05 SD` case is included because democratic trust outcomes may be harder
to move than immediate climate-action intentions.

## Setup

The script uses DeclareDesign, fabricatr, randomizr, estimatr, marginaleffects,
dplyr, and ggplot2. The seed is fixed for reproducibility.

```{r}
#| label: setup
#| echo: true
#| output: false
library(DeclareDesign)
library(fabricatr)
library(randomizr)
library(dplyr)
library(ggplot2)

set.seed(20260413)
```

## Design Knobs

The panel version of the example assumes a baseline-to-follow-up correlation.
Higher stability means baseline adjustment gives more precision. The attrition
grid is included because even a short panel can lose respondents between waves.

```{r}
#| label: design-knobs
#| cache: false
#| output: false
# Planned total panel sample size (wave 1)
N <- 4000

# Default treatment effect for single-design runs; grid used in diagnosis
tau <- 0.10
tau_grid <- c(0.05, 0.08, 0.10, 0.15, 0.20)  # SD units (standardised outcome)

# Test-retest correlation: Pearson r between wave-1 and wave-2 potential outcome
# under no treatment. Bounded [-1, 1]; plausible range for attitude outcomes ~0.2-0.6.
# Determines how much variance the baseline covariate absorbs in ANCOVA.
rho_y <- 0.40
rho_grid <- c(0.20, 0.40, 0.60)

# Sample sizes for the N-sensitivity analysis
N_grid <- c(1000, 2000, 4000)

# Probability of non-response at wave 2 (non-differential by default)
attrition_rate <- 0.30
# Additional attrition probability for treated units (0 = no differential attrition)
differential_attrition <- 0.00

# Number of simulations per design cell. Raise to 500+ for a final run.
sims <- 500
```

## Helper Functions

The latent outcome is mapped to a simple 1-5 Likert-style scale so the same
design can be inspected on both a standardized and an ordinal-looking metric.

```{r}
#| label: helpers
#| cache: false
#| output: false
latent_to_likert <- function(x) {
  as.integer(cut(
    x,
    breaks = c(-Inf, -1.2, -0.35, 0.35, 1.2, Inf),
    labels = FALSE
  ))
}

diagnosands_power <- declare_diagnosands(
  power = mean(p.value < 0.05, na.rm = TRUE),
  bias = mean(estimate - estimand, na.rm = TRUE),
  rmse = sqrt(mean((estimate - estimand)^2, na.rm = TRUE)),
  coverage = mean(
    conf.low <= estimand & conf.high >= estimand,
    na.rm = TRUE
  ),
  mean_estimate = mean(estimate, na.rm = TRUE),
  mean_estimand = mean(estimand, na.rm = TRUE),
  mean_se = mean(std.error, na.rm = TRUE),
  type_s_error = mean(
    p.value < 0.05 & sign(estimate) != sign(estimand),
    na.rm = TRUE
  )
)

```

## Data-Generating Model

The simulated population includes basic demographic variation, ideology, and a
binary climate concern indicator. The untreated follow-up outcome is correlated
with baseline outcome, and treatment shifts the follow-up potential outcome.

```{r}
#| label: model
#| cache: false
#| output: false
model_wp2 <- declare_model(
  N = N,
  gender = factor(
    sample(c("Woman", "Man"), N, replace = TRUE, prob = c(0.51, 0.49))
  ),
  age_group = factor(
    sample(
      c("18-34", "35-54", "55+"),
      N,
      replace = TRUE,
      prob = c(0.28, 0.35, 0.37)
    )
  ),
  education = factor(
    sample(
      c("Low", "Medium", "High"),
      N,
      replace = TRUE,
      prob = c(0.25, 0.45, 0.30)
    )
  ),
  ideology3 = factor(
    sample(
      c("Left", "Center", "Right"),
      N,
      replace = TRUE,
      prob = c(0.34, 0.33, 0.33)
    )
  ),
  climate_high = rbinom(
    N,
    1,
    ifelse(ideology3 == "Left", 0.72, ifelse(ideology3 == "Center", 0.55, 0.38))
  ),
  climate_concern_high = factor(
    ifelse(climate_high == 1, "Higher concern", "Lower concern")
  ),
  baseline_mu =
    0.20 * (gender == "Woman") +
    0.15 * (age_group == "55+") +
    0.10 * (education == "High") +
    0.35 * (ideology3 == "Left") -
    0.25 * (ideology3 == "Right") +
    0.45 * climate_high,
  Y_pre_raw = baseline_mu + rnorm(N),
  Y_pre_latent = as.numeric(scale(Y_pre_raw)),
  post_shock = rnorm(N),
  Y0_latent = rho_y * Y_pre_latent + sqrt(1 - rho_y^2) * post_shock,
  tau_i = tau +
    0.03 * climate_high -
    0.02 * (ideology3 == "Right"),
  Y_latent_Z_0 = Y0_latent,
  Y_latent_Z_1 = Y0_latent + tau_i,
  Y_1_5_Z_0 = latent_to_likert(Y_latent_Z_0),
  Y_1_5_Z_1 = latent_to_likert(Y_latent_Z_1)
)
```

## Inquiries, Assignment, and Measurement

The estimands are the average treatment effects on the latent scale and on the
coarsened 1-5 scale. Treatment is assigned 50/50. The measurement step reveals
the realized outcomes.

```{r}
#| label: inquiries-assignment-measurement
#| output: false
inquiries_wp2 <- declare_inquiry(
  ATE_latent = mean(Y_latent_Z_1 - Y_latent_Z_0),
  ATE_1_5 = mean(Y_1_5_Z_1 - Y_1_5_Z_0)
)

assignment_wp2 <- declare_assignment(
  Z = complete_ra(N = N, prob = 0.5)
)

measurement_wp2 <- declare_measurement(
  Y_latent = reveal_outcomes(Y_latent ~ Z),
  Y_1_5 = reveal_outcomes(Y_1_5 ~ Z)
)
```

## Attrition

The attrition block simulates wave-2 response probability and then filters the
analysis sample to respondents only.

```{r}
#| label: attrition
#| cache: false
#| output: false
attrition_wp2 <- declare_measurement(
  responds_post = rbinom(
    N,
    1,
    plogis(
      qlogis(1 - attrition_rate) -
        differential_attrition * Z +
        0.15 * climate_high -
        0.10 * (education == "Low")
    )
  ),
  filter = responds_post == 1
)
```

## Estimators

The example compares a post-only estimator with a baseline-adjusted panel
estimator. Both use `lm` (OLS, IID standard errors).

```{r}
#| label: estimators
#| cache: false
#| output: false
post_only_estimators <-
  declare_estimator(
    Y_latent ~ Z,
    .method = lm,
    inquiry = "ATE_latent",
    term = "Z",
    label = "post_only_latent"
  ) +
  declare_estimator(
    Y_1_5 ~ Z,
    .method = lm,
    inquiry = "ATE_1_5",
    term = "Z",
    label = "post_only_1_5"
  )

panel_estimators <-
  declare_estimator(
    Y_latent ~ Z + Y_pre_latent,
    .method = lm,
    inquiry = "ATE_latent",
    term = "Z",
    label = "baseline_adjusted_latent"
  ) +
  declare_estimator(
    Y_1_5 ~ Z + Y_pre_latent,
    .method = lm,
    inquiry = "ATE_1_5",
    term = "Z",
    label = "baseline_adjusted_1_5"
  )
```

## Designs

Two designs are compared: post-only and a baseline-adjusted panel with 20%
attrition.

```{r}
#| label: designs
#| cache: false
#| output: false
design_post_only <-
  model_wp2 +
  inquiries_wp2 +
  assignment_wp2 +
  measurement_wp2 +
  post_only_estimators

design_panel <-
  model_wp2 +
  inquiries_wp2 +
  assignment_wp2 +
  measurement_wp2 +
  attrition_wp2 +
  panel_estimators
```

## Diagnosis

The `redesign()` calls vary the treatment effect, outcome stability, and sample
size across both designs. The diagnosis table is the main output for a power
assessment.

```{r}
#| label: diagnosis
#| cache: false
post_only_grid <- redesign(
  design_post_only,
  tau = tau_grid,
  rho_y = rho_grid,
  N = N_grid
)

panel_grid <- redesign(
  design_panel,
  tau = tau_grid,
  rho_y = rho_grid,
  N = N_grid
)

diagnosis_post_only <- diagnose_designs(
  post_only_grid,
  diagnosands = diagnosands_power,
  sims = sims,
  bootstrap_sims = FALSE
)

diagnosis_panel <- diagnose_designs(
  panel_grid,
  diagnosands = diagnosands_power,
  sims = sims,
  bootstrap_sims = FALSE
)

diagnosis_results <- bind_rows(
  diagnosis_post_only$diagnosands_df |>
    mutate(scenario = "post_only"),
  diagnosis_panel$diagnosands_df |>
    mutate(scenario = "panel_with_attrition")
)

```

## Plots

All plots focus on the latent standardized outcome.

```{r}
#| label: plot-power
#| echo: false
#| fig-cap: "Statistical power by true effect size (SD units) and sample size, at rho_y = 0.40. Dashed line marks 80% power. Panel design includes 30% attrition and baseline covariate adjustment."
latent_results <- diagnosis_results |>
  filter(grepl("latent", estimator))

main_results <- latent_results |>
  filter(rho_y == 0.40)

power_plot <- main_results |>
  ggplot(aes(x = tau, y = power, color = estimator, group = estimator)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point() +
  facet_wrap(~ N, labeller = label_both) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "True treatment effect, SD units",
    y = "Power",
    color = "Estimator"
  ) +
  theme_minimal()

print(power_plot)
```

```{r}
#| label: plot-coverage
#| echo: false
#| fig-cap: "95% CI coverage by true effect size and sample size, at rho_y = 0.40. Dashed line marks the nominal 0.95 level. Sustained deviations indicate SE miscalibration."
coverage_plot <- main_results |>
  ggplot(aes(x = tau, y = coverage, color = estimator, group = estimator)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point() +
  facet_wrap(~ N, labeller = label_both) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "True treatment effect, SD units",
    y = "Coverage (nominal 95%)",
    color = "Estimator"
  ) +
  theme_minimal()

print(coverage_plot)
```

```{r}
#| label: plot-bias
#| echo: false
#| fig-cap: "Bias (mean estimate minus estimand) by true effect size and sample size, at rho_y = 0.40. Dashed line at zero. Non-zero bias in a randomised design signals a DGP or estimator misspecification."
bias_plot <- main_results |>
  ggplot(aes(x = tau, y = bias, color = estimator, group = estimator)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line() +
  geom_point() +
  facet_wrap(~ N, labeller = label_both) +
  labs(
    x = "True treatment effect, SD units",
    y = "Bias",
    color = "Estimator"
  ) +
  theme_minimal()

print(bias_plot)
```

## Interpretation

```{r}
#| label: results
#| output: false
target_tau <- 0.10
target_rho <- 0.40

# Power and coverage at the main planning assumption (N = 4000)
target_N <- 4000

key <- diagnosis_results |>
  filter(grepl("latent", estimator), tau == target_tau, rho_y == target_rho,
         N == target_N)

po_power    <- key |> filter(scenario == "post_only")            |> pull(power)    |> round(2)
panel_power <- key |> filter(scenario == "panel_with_attrition") |> pull(power)    |> round(2)
po_cov      <- key |> filter(scenario == "post_only")            |> pull(coverage) |> round(2)
panel_cov   <- key |> filter(scenario == "panel_with_attrition") |> pull(coverage) |> round(2)

# MDE: smallest tau reaching 80% power at rho = 0.60, N = 4000
mde_at <- function(sc) {
  v <- diagnosis_results |>
    filter(grepl("latent", estimator), scenario == sc, rho_y == target_rho,
           N == target_N, power >= 0.80) |>
    arrange(tau) |>
    slice(1) |>
    pull(tau)
  if (length(v) == 0) paste0(">", max(tau_grid)) else as.character(v)
}

mde_po    <- mde_at("post_only")
mde_panel <- mde_at("panel_with_attrition")

# Power gain from panel at target effect
power_gain <- round(panel_power - po_power, 2)

# Breakeven rho: panel beats post-only iff rho > sqrt(attrition_rate)
breakeven_rho <- round(sqrt(attrition_rate), 2)

# Coverage deviation from nominal
po_cov_dev    <- round(abs(po_cov    - 0.95), 2)
panel_cov_dev <- round(abs(panel_cov - 0.95), 2)
```

This is an example design, not a final power analysis. Numbers reported here
update automatically when `sims` is increased.

**Effect-size anchor.** The main planning assumption is `tau = 0.10 SD` with
outcome stability `rho_y = 0.40` and N = 4000. At these values, the post-only
design reaches power `r po_power` and the baseline-adjusted panel (with 30%
attrition) reaches power `r panel_power` — a gain of `r power_gain` percentage
points from adjusting for the pre-test.

**Minimum detectable effect.** At `rho_y = 0.60`, the smallest effect where
power first exceeds 80% is `r mde_po` SD for the post-only design and
`r mde_panel` SD for the panel. The panel MDE is lower because the
pre-test covariate absorbs outcome variance.

**Sensitivity to outcome stability.** The `rho_y` grid (0.20, 0.40, 0.60)
spans the range from weakly to moderately correlated waves. The power advantage
of the panel over post-only grows with `rho_y`: when pre-post correlation is
high, the baseline covariate explains more residual variance and the panel
gains more precision.

**Attrition and breakeven.** The panel design builds in 30% attrition with
modest educational and climate-concern gradients. Attrition reduces effective N
but does not induce bias here (`differential_attrition = 0`). The panel design
gains efficiency over post-only only when `rho_y > sqrt(attrition_rate)`,
i.e., `rho_y >` `r breakeven_rho`. Below this threshold the sample loss from
attrition outweighs the variance reduction from the baseline covariate. With
30% attrition the breakeven is `r breakeven_rho`, meaning the panel advantage
is modest at `rho_y = 0.60` and absent at `rho_y = 0.20` and `rho_y = 0.40`. Differential
attrition would require separate sensitivity
analysis.

**Coverage.** At the target effect and `rho_y = 0.60`, 95% CI coverage is
`r po_cov` (post-only) and `r panel_cov` (panel), deviating from nominal
by `r po_cov_dev` and `r panel_cov_dev` respectively. Deviations larger than
0.03–0.05 would warrant investigation of SE calibration.

## Running

Install the needed R packages:

```{r}
#| label: install
#| eval: false
install.packages(c(
  "DeclareDesign", "fabricatr", "randomizr",
  "dplyr", "ggplot2"
))
```

Then render the document or source the code in an R environment that has
DeclareDesign installed.

## Sources

- `cad_short.pdf`, WP2 section.
- `Ballew et al. - 2022 - Changing minds about global warming vicarious exp.pdf`
- `Sinclair et al. - 2024 - Behavioral Interventions Motivate Action to Address Climate Change.pdf`
- `Goldwert et al. - 2026 - A megastudy of behavioral interventions to catalyze public, political, and financial climate advocac.pdf`
- DeclareDesign documentation: <https://declaredesign.org/r/declaredesign/>
- `declare_estimator()` documentation:
  <https://declaredesign.org/r/declaredesign/reference/declare_estimator.html>

